---
title: "MTHM506 - Statistical Data Modelling"
author: "Joshua Harrison"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document:
    css: style.css
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r nlmodel, include=FALSE}
load("C:/Users/jh1376/OneDrive - University of Exeter/MSc Applied Data Science and Statistics/MTHM506 - Statistical Data Modelling/Coursework 1/datasets.RData")
library(ggplot2)
library(rmarkdown)
library(tidyverse)
library(lmerTest)
```
# **Question 2**

The dataset `aids` contains quarterly AIDS cases (`y_i`) from January 1983 to March 1994. We analyze two competing models:

1. **Poisson Model (Model 1)**:
   \[
   Y_i \sim Pois(\lambda_i)
   \]
   \[
   \log(\lambda_i) = \beta_0 + \beta_1 x_i
   \]
   
2. **Normal Model (Model 2)**:
   \[
   Y_i \sim N(\mu_i, \sigma^2)
   \]
   \[
   \log(\mu_i) = \gamma_0 + \gamma_1 x_i
   \]


**(a) Plot yi against xi and comment on whether the two proposed models are sensible in terms on the distribution and the relationship of x with the mean**


# Scatter plot of cases vs time
ggplot(aids, aes(x = date, y = cases)) +
  geom_point(color = "blue") +
  labs(title = "Quarterly AIDS Cases in the UK",
       x = "Time (Quarters since 1983)", y = "Number of Cases") +
  theme_minimal()

ednpca

